\chapter{Inverse Functions and Limits}

% expansion:intuition — chapter overview synthesising the manuscript's two themes
This chapter develops two ideas that together form the starting point of calculus. The first is the inverse of a function: when a function sends each input to a distinct output, we can ask how to recover the input from the output. This leads to one-to-one functions, inverse functions, and the inverse trigonometric functions. The second is the limit of a function. Many quantities of interest in calculus --- the instantaneous rate of change of a moving object, the slope of a tangent line, the behavior of a function near a hole in its graph --- depend not on the value of the function at a single point but on its behavior as the input approaches that point. We study limits first informally through tables and graphs, then through algebraic laws that let us compute them systematically, and finally through the precise \(\varepsilon\)-\(\delta\) definition that places every preceding argument on a rigorous footing.

% expansion:summary — learning outcomes distilled from the manuscript's coverage
\paragraph{By the end of this chapter, you will be able to:}
\begin{itemize}
    \item decide when a function is one-to-one and find its inverse when it exists;
    \item work with the inverse trigonometric functions \(\arcsin\), \(\arccos\), \(\arctan\), and their companions, together with their principal intervals;
    \item estimate the value of a limit from tables, graphs, and algebraic simplification, and handle one-sided and infinite limits;
    \item compute limits using the limit laws, factoring and rationalisation, and the squeeze theorem;
    \item state and apply the precise \(\varepsilon\)-\(\delta\) definition of a limit to verify a given limit rigorously.
\end{itemize}

\section{Inverse Functions}

% expansion:intuition — section opener previewing this section's structure, pointing back to the chapter overview rather than restating its framing
Recall that a function \(f\) assigns a value to every input in its domain. The reverse machinery --- how to recover the input from the output --- was named in the chapter overview; this section works it out. We proceed in three stages: identify which functions admit inversion (the one-to-one property), define the inverse function for those that do, and give an explicit procedure for constructing it. Before the formal definitions, two simple examples show when inversion is possible and when it is not.

% expansion:intuition — two numeric warm-ups tracking the manuscript's opening "Ex" pair
\begin{enumerate}
    \item \(f(x) = x\) on \([0, 1]\). Here \(f(0) = 0\), \(f(\tfrac{1}{2}) = \tfrac{1}{2}\), \(f(\tfrac{1}{8}) = \tfrac{1}{8}\); each input produces its own distinct output.
    \item \(f(x) = x^{2}\) on \([-1, 1]\). Here \(f(0) = 0\), \(f(\tfrac{1}{2}) = \tfrac{1}{4}\), and \(f(-\tfrac{1}{2}) = \tfrac{1}{4}\); two different inputs produce the same output.
\end{enumerate}

The difference between these two cases is the central point. In the first, every output corresponds to exactly one input. In the second, the output \(\tfrac{1}{4}\) corresponds to both \(\tfrac{1}{2}\) and \(-\tfrac{1}{2}\), so there is no unambiguous way to go backwards from output to input.

\subsection{One-to-one functions}

A function can be reversed only when each output comes from exactly one input. This motivates the following definition.

\begin{definition}
Let \(f\) be a function with domain \(A\). We say that \(f\) is \emph{one-to-one}\index{one-to-one function} if
\[
f(x_{1}) \neq f(x_{2}) \qquad \text{whenever } x_{1} \neq x_{2}.
\]
Equivalently, \(f\) is one-to-one if
\[
f(x_{1}) = f(x_{2}) \implies x_{1} = x_{2}.
\]

% expansion:intuition — Informally gloss (SPEC §3 allows this inside definition)
\emph{Informally, a function is one-to-one when different inputs always produce different outputs.}
\end{definition}

% expansion:example — student-ID illustration, following the manuscript's explicit hint "(use example) 學考試號 number → students"
\begin{workedexample}
\begin{example}
A class of students is assigned identification numbers, arranged so that no two students share a number. Define
\[
f(\text{student}) = \text{the student's identification number}.
\]
Is \(f\) one-to-one?
\end{example}

\begin{solution}
By construction, no two students share a number, so distinct students always produce distinct identification numbers. Therefore \(f\) is one-to-one.
\end{solution}
\end{workedexample}

Now we return to the two mathematical examples from the opening.

\begin{workedexample}
\begin{example}
Determine which of the following functions are one-to-one on their given domains:
\begin{conditiondisplay}
f(x) &= x, \qquad & x \in [0, 1],\\
g(x) &= x^{2}, \qquad & x \in [-1, 1].
\end{conditiondisplay}
\end{example}

\begin{solution}
The function \(f(x) = x\) is one-to-one on \([0, 1]\) because each input is sent to itself, so distinct inputs always give distinct outputs.

The function \(g(x) = x^{2}\) is not one-to-one on \([-1, 1]\) because
\begin{aligneddisplay}
g\!\left(\tfrac{1}{2}\right) &= \left(\tfrac{1}{2}\right)^{2} = \tfrac{1}{4},\\
g\!\left(-\tfrac{1}{2}\right) &= \left(-\tfrac{1}{2}\right)^{2} = \tfrac{1}{4}.
\end{aligneddisplay}
Two different inputs produce the same output, so \(g\) is not one-to-one.
\end{solution}
\end{workedexample}

The function \(g(x) = x^{2}\) on \([-1, 1]\) illustrates why not every function can be reversed: the value \(\tfrac{1}{4}\) is produced by both \(\tfrac{1}{2}\) and \(-\tfrac{1}{2}\), so no single rule can take \(\tfrac{1}{4}\) back to a uniquely determined input.

Looking at the graph of a function gives a quick way to decide whether it is one-to-one.

\begin{remark}[Horizontal line test]\index{horizontal line test}
A horizontal line \(y = c\) meets the graph of \(f\) at a point \((x, c)\) exactly when \(f(x) = c\). Therefore, \emph{a function is one-to-one if and only if no horizontal line intersects its graph more than once}. If the line \(y = c\) meets the graph at two distinct points \((x_{1}, c)\) and \((x_{2}, c)\), then \(f(x_{1}) = c = f(x_{2})\) with \(x_{1} \neq x_{2}\), so \(f\) fails to be one-to-one; conversely, if every horizontal line meets the graph at most once, each output corresponds to at most one input. \Cref{fig:horizontal-line-test} shows the two situations.
\end{remark}

\begin{figure}[H]
\centering
\begin{minipage}{0.46\textwidth}
\centering
\begin{tikzpicture}[scale=0.9]
\draw[->] (-2.3,0) -- (2.1,0) node[right] {$x$};
\draw[->] (0,-2.1) -- (0,2.2) node[above] {$y$};
\draw[blue, thick] (-1.5,-1.5) -- (1.3,1.3);
\draw[red, dashed] (-2.1,1) -- (1.3,1);
\fill (1,1) circle (2pt);
\node[blue] at (1.4,1.55) {$f_{1}(x)$};
\node[below right] at (0.85,1) {$(x_{1},c)$};
\node[red, right] at (1.35,1) {$y = c$};
\end{tikzpicture}

\vspace{0.4em}
One-to-one.\\
Line meets the graph at most once.
\end{minipage}
\hfill
\begin{minipage}{0.46\textwidth}
\centering
\begin{tikzpicture}[scale=0.9]
\draw[->] (-2.3,0) -- (2.1,0) node[right] {$x$};
\draw[->] (0,-2.1) -- (0,2.2) node[above] {$y$};
\draw[blue, thick, domain=-1.35:1.35, samples=100] plot (\x,{1.15*\x*\x});
\draw[red, dashed] (-1.85,1.25) -- (1.85,1.25);
\fill (-1.04,1.25) circle (2pt);
\fill (1.04,1.25) circle (2pt);
\node[blue] at (1.5,2.15) {$f_{2}(x)$};
\node[above left] at (-1.2,1.25) {$(x_{1},c)$};
\node[above right] at (1.2,1.25) {$(x_{2},c)$};
\node[red, right] at (1.9,1.25) {$y = c$};
\end{tikzpicture}

\vspace{0.4em}
Not one-to-one.\\
Line meets the graph more than once.
\end{minipage}
\caption{Geometric interpretation of the horizontal line test.}
\label{fig:horizontal-line-test}
\end{figure}

\subsection{Inverse functions}

If \(f\) is one-to-one, then we can define a second function that undoes \(f\): given an output of \(f\), it returns the unique input that produced it.

\begin{definition}
Let \(f\) be a one-to-one function with domain \(A\) and range \(B\). The \emph{inverse function}\index{inverse function}\index{inverse function@$f^{-1}$} of \(f\), denoted by \(f^{-1}\), has domain \(B\) and range \(A\), and is defined by
\[
f^{-1}(y) = x \quad \Longleftrightarrow \quad f(x) = y \ \text{for any } y \in B.
\]
\end{definition}

\begin{remark}
When we focus on the inverse function itself, it is often convenient to use \(x\) as its independent variable. In that case, we write
\[
f^{-1}(x) = y \quad \Longleftrightarrow \quad f(y) = x.
\]
\end{remark}

\begin{theorem}
A function has an inverse if and only if it is one-to-one.
\end{theorem}

% expansion:formula — short proof from the definition; short, standard, illustrative
\begin{proof}
Suppose first that \(f\) has an inverse \(f^{-1}\). If \(f(x_{1}) = f(x_{2})\), applying \(f^{-1}\) to both sides gives
\[
f^{-1}\!\bigl(f(x_{1})\bigr) = f^{-1}\!\bigl(f(x_{2})\bigr),
\]
so \(x_{1} = x_{2}\). Therefore \(f\) is one-to-one. Conversely, suppose \(f\) is one-to-one with domain \(A\) and range \(B\). For each \(y \in B\), there exists some \(x \in A\) with \(f(x) = y\), and the one-to-one property forces this \(x\) to be unique. The correspondence \(y \mapsto x\) therefore defines an inverse function \(f^{-1} : B \to A\).
\end{proof}

\begin{workedexample}
\begin{example}
Find the inverse of each function.
\begin{enumerate}
    \item \(f(x) = x\).
    \item \(g(x) = x^{3}\).
\end{enumerate}
\end{example}

\begin{solution}
\solutionbreak
\begin{enumerate}
    \item Since \(f(x) = x\) leaves every input unchanged, \(f\) is its own inverse, so \(f^{-1}(x) = x\).

    \item To find the inverse of \(g\), let \(y = x^{3}\). Solving for \(x\) gives \(x = \sqrt[3]{y}\). Interchanging the names of the variables, \(g^{-1}(x) = \sqrt[3]{x}\). As a check,
    \[
    g\!\bigl(g^{-1}(x)\bigr) = \bigl(\sqrt[3]{x}\bigr)^{3} = x.
    \]
\end{enumerate}
\end{solution}
\end{workedexample}

The defining relation \(f^{-1}(y) = x \Longleftrightarrow f(x) = y\) yields two composition identities.

\begin{proposition}
Let \(f\) be a one-to-one function with domain \(A\) and range \(B\). Then
\pairdisplay{f^{-1}\!\bigl(f(x)\bigr) = x,\quad \text{for every } x \in A,}{f\!\bigl(f^{-1}(y)\bigr) = y,\quad \text{for every } y \in B.}
\end{proposition}

\Cref{fig:inverse-composition} illustrates both identities: \(f\) and \(f^{-1}\) travel between \(A\) and \(B\) in opposite directions, and composing them recovers the original input.

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[
        scale=0.95,
        transform shape,
        >=Stealth,
        every node/.style={font=\small},
        lab/.style={midway, fill=white, inner sep=1pt},
        mainnode/.style={text=blue!50!black},
        fwd/.style={->, very thick, draw=blue!50!blue},
        bwd/.style={->, very thick, draw=red!70!blue}
    ]

    \node[mainnode] (xA) at (0,0) {$x\in A$};
    \node[mainnode] (fxB) at (4.8,0) {$f(x)\in B$};

    \draw[fwd, bend left=20] (xA) to node[lab, above, text=blue!40!black] {$f$} (fxB);
    \draw[bwd, bend left=20] (fxB) to node[lab, below, text=red!70!black] {$f^{-1}$} (xA);

    \node[text=blue!60!black] at (2.4,-1.5) {$f^{-1}(f(x))=x$};

    \node[mainnode] (yB) at (8.2,0) {$y\in B$};
    \node[mainnode] (finvyA) at (13,0) {$f^{-1}(y)\in A$};

    \draw[bwd, bend left=20] (yB) to node[lab, above, text=red!70!black] {$f^{-1}$} (finvyA);
    \draw[fwd, bend left=20] (finvyA) to node[lab, below, text=blue!40!black] {$f$} (yB);

    \node[text=blue!60!black] at (10.6,-1.5) {$f\bigl(f^{-1}(y)\bigr)=y$};

    \end{tikzpicture}
    \caption{A one-to-one function and its inverse.}
    \label{fig:inverse-composition}
\end{figure}

% expansion:formula — standard reflection-across-y=x consequence
Graphically, if the graph of \(f^{-1}\) is drawn in the same coordinate plane as the graph of \(f\), the two graphs are reflections of one another across the line \(y = x\). The point \((a, b)\) lies on the graph of \(f\) exactly when \((b, a)\) lies on the graph of \(f^{-1}\).

\subsection{Finding the inverse of a function}

When \(f\) is one-to-one, we can usually construct its inverse by a short procedure.

\begin{strategy}[Finding the inverse of a one-to-one function]
\begin{enumerate}
    \item Write \(y = f(x)\).
    \item Solve this equation for \(x\) in terms of \(y\).
    \item Interchange the names \(x\) and \(y\). The resulting equation gives \(y = f^{-1}(x)\).
\end{enumerate}
\end{strategy}

\begin{workedexample}
\begin{example}
Find the inverse of \(f(x) = x^{3} + 2\).
\end{example}

\begin{solution}
Let \(y = x^{3} + 2\). Solving for \(x\) gives
\begin{aligneddisplay}
x^{3} &= y - 2,\\
x &= \sqrt[3]{y - 2}.
\end{aligneddisplay}
Interchanging the variables, \(y = \sqrt[3]{x - 2}\), so \(f^{-1}(x) = \sqrt[3]{x - 2}\). As a check,
\begin{aligneddisplay}
f\!\bigl(f^{-1}(x)\bigr)
&= \left(\sqrt[3]{x - 2}\right)^{3} + 2\\
&= (x - 2) + 2\\
&= x.
\end{aligneddisplay}
\end{solution}
\end{workedexample}

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.1, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

\section{Inverse Trigonometric Functions}

% expansion:application — ties inverse trig to problems from geometry and physics that motivate them
The trigonometric functions are not one-to-one on all of \(\mathbb{R}\), so they do not have inverse functions on their full domains. Inverse trigonometric functions arise whenever we know a ratio and want to recover an angle: from the slope of a ramp, we get an angle of inclination via \(\arctan\); from a ratio of sides in a right triangle, we get an angle via \(\arcsin\) or \(\arccos\). To build these inverses, we must first restrict the domain of each trigonometric function to an interval on which it is one-to-one.

Consider the sine function on \(\mathbb{R}\). As \cref{fig:sine-not-one-to-one} shows, a horizontal line can intersect the graph of \(y = \sin x\) more than once, so \(\sin\) is not one-to-one on \(\mathbb{R}\).

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=11.5cm,
height=4cm,
axis lines=middle,
xmin=-7,
xmax=7,
ymin=-1.3,
ymax=1.3,
xtick={-6.283185307,-3.141592654,0,3.141592654,6.283185307},
xticklabels={\(-2\pi\),\(-\pi\),\(0\),\(\pi\),\(2\pi\)},
ytick={-1,1},
yticklabels={\(-1\),\(1\)},
xlabel={\(x\)},
xlabel style={at={(axis description cs:1.0,0.48)}, anchor=west},
ylabel={\(y\)},
samples=300,
clip=false
]
\addplot[thick, blue, domain=-6.8:6.8] {sin(deg(x))};
\draw[red, dashed] (axis cs:-6.8,0.5) -- (axis cs:6.8,0.5);
\fill (axis cs:0.523598776,0.5) circle[radius=1.1pt];
\fill (axis cs:2.617993878,0.5) circle[radius=1.1pt];
\fill (axis cs:6.806783967,0.5) circle[radius=1.1pt];
\fill (axis cs:-3.665191429,0.5) circle[radius=1.1pt];
\fill (axis cs:-5.759586531,0.5) circle[radius=1.1pt];
\node at (axis cs:3.3,0.95) {\(\sin x\)};
\node[right] at (axis cs:7.15,0.5) {\(y=c\)};
\end{axis}
\end{tikzpicture}
\caption{The sine function on \(\mathbb{R}\) is not one-to-one.}
\label{fig:sine-not-one-to-one}
\end{figure}

\subsection{The inverse sine function}

To define the inverse sine, we restrict \(\sin x\) to the interval
\[
-\tfrac{\pi}{2} \le x \le \tfrac{\pi}{2}.
\]
On this interval, \(\sin x\) is one-to-one and its values cover the range \([-1, 1]\). The restricted branch is shown in \cref{fig:restricted-sine}.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=8.5cm,
height=5.6cm,
axis lines=middle,
xmin=-1.9,
xmax=1.9,
ymin=-1.2,
ymax=1.2,
xtick={-1.570796327,1.570796327},
xticklabels={\(-\tfrac{\pi}{2}\),\(\tfrac{\pi}{2}\)},
ytick={-1,1},
yticklabels={\(-1\),\(1\)},
xlabel={\(x\)},
ylabel={\(y\)},
samples=200,
clip=false
]
\addplot[thick,blue, domain=-1.570796327:1.570796327] {sin(deg(x))};
\node at (axis cs:0.95,1.1) {\(\sin x\)};
\end{axis}
\end{tikzpicture}
\caption{Restricted branch of the sine function used to define \(\arcsin x\).}
\label{fig:restricted-sine}
\end{figure}

\begin{definition}
The \emph{inverse sine function}\index{inverse sine function}\index{arcsine@$\arcsin$}, denoted \(\arcsin x\) or \(\sin^{-1} x\), is defined by
\[
\arcsin x = y \quad \Longleftrightarrow \quad \sin y = x \text{ and } -\tfrac{\pi}{2} \le y \le \tfrac{\pi}{2}.
\]
Its domain is \([-1, 1]\) and its range is \(\left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\).
\end{definition}

% expansion:caution — notation trap the manuscript uses both spellings but does not warn about
\begin{caution}\index{sine inverse vs reciprocal@$\sin^{-1}$ vs $1/\sin$}
The notation \(\sin^{-1} x\) is also used for \(\arcsin x\). It does \emph{not} mean the reciprocal \(\bigl(\sin x\bigr)^{-1} = 1 / \sin x\). When reading a source that uses \(\sin^{-1}\), check the context to decide which is meant.
\end{caution}

The composition identities for \(\arcsin\) take the following form.

\begin{proposition}[Inverse sine identities]
\begin{conditiondisplay}
\sin(\arcsin x) &= x, \qquad & -1 \le x \le 1,\\
\arcsin(\sin x) &= x, \qquad & -\tfrac{\pi}{2} \le x \le \tfrac{\pi}{2}.
\end{conditiondisplay}
\end{proposition}

% expansion:caution — the manuscript gives the identity with its restriction but does not flag what goes wrong outside the principal interval
\begin{caution}\index{arcsin of sin@$\arcsin(\sin x)$ outside principal interval}
The identity \(\arcsin(\sin x) = x\) holds only when \(x \in \left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\). Outside this interval, the right-hand side is replaced by the unique value in \(\left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\) with the same sine; for instance,
\[
\arcsin(\sin \pi) = \arcsin(0) = 0 \neq \pi.
\]
\end{caution}

\begin{workedexample}
\begin{example}
Evaluate the following expressions.
\begin{enumerate}
    \item \(\arcsin\!\left(\tfrac{1}{2}\right)\).
    \item \(\tan\!\bigl(\arcsin\!\left(\tfrac{1}{3}\right)\bigr)\).
\end{enumerate}
\end{example}

\begin{solution}
\solutionbreak
\begin{enumerate}
    \item We need the angle \(y \in \left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\) with \(\sin y = \tfrac{1}{2}\). Since \(\sin(\pi/6) = \tfrac{1}{2}\), we have \(\arcsin\!\left(\tfrac{1}{2}\right) = \tfrac{\pi}{6}\).

    \item Let \(\theta = \arcsin\!\left(\tfrac{1}{3}\right)\), so \(\sin\theta = \tfrac{1}{3}\) with \(\theta \in \left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\). Because \(\sin\theta > 0\), the angle \(\theta\) lies in the first quadrant, and in a right triangle with hypotenuse \(3\) and opposite side \(1\), the adjacent side is
    \[
    a = \sqrt{3^{2} - 1^{2}} = \sqrt{8} = 2\sqrt{2}.
    \]

    \begin{center}
    \begin{tikzpicture}[scale=1.1]
    \coordinate (A) at (0,0);
    \coordinate (B) at (3,0);
    \coordinate (C) at (3,1);
    \draw[thick] (A) -- (B) -- (C) -- cycle;
    \draw (2.75,0) -- (2.75,0.25) -- (3,0.25);
    \node[left] at (1.2,0.18) {\(\theta\)};
    \node[below] at (1.5,0) {\(2\sqrt{2}\)};
    \node[right] at (3,0.5) {\(1\)};
    \node[above left] at (1.45,0.5) {\(3\)};
    \end{tikzpicture}
    \end{center}

    Therefore \(\tan\!\bigl(\arcsin\!\left(\tfrac{1}{3}\right)\bigr) = \tan\theta = \dfrac{1}{2\sqrt{2}}\).
\end{enumerate}
\end{solution}
\end{workedexample}

\subsection{The inverse cosine function}

The cosine function is one-to-one on the interval \(0 \le x \le \pi\), shown in \cref{fig:restricted-cosine}.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=8.5cm,
height=3.8cm,
axis lines=middle,
xmin=-0.3,
xmax=3.5,
ymin=-1.3,
ymax=1.7,
xtick={0,3.141592654},
xticklabels={\(0\),\(\pi\)},
ytick={-1,1},
yticklabels={\(-1\),\(1\)},
xlabel={\(x\)},
ylabel={\(y\)},
samples=200,
clip=false
]
\addplot[thick,blue, domain=0:3.141592654] {cos(deg(x))};
\node at (axis cs:1.2,0.72) {\(\cos x\)};
\end{axis}
\end{tikzpicture}
\caption{Restricted branch of the cosine function used to define \(\arccos x\).}
\label{fig:restricted-cosine}
\end{figure}

\begin{definition}
The \emph{inverse cosine function}\index{inverse cosine function}\index{arccosine@$\arccos$}, denoted \(\arccos x\) or \(\cos^{-1} x\), is defined by
\[
\arccos x = y \quad \Longleftrightarrow \quad \cos y = x \text{ and } 0 \le y \le \pi.
\]
Its domain is \([-1, 1]\) and its range is \([0, \pi]\).
\end{definition}

\begin{proposition}[Inverse cosine identities]
\begin{conditiondisplay}
\cos(\arccos x) &= x, \qquad & -1 \le x \le 1,\\
\arccos(\cos x) &= x, \qquad & 0 \le x \le \pi.
\end{conditiondisplay}
\end{proposition}

\subsection{The inverse tangent function}

The tangent function is one-to-one on the interval \(-\tfrac{\pi}{2} < x < \tfrac{\pi}{2}\), shown in \cref{fig:restricted-tangent}.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=8.5cm,
height=8.5cm,
axis lines=middle,
xmin=-1.9,
xmax=1.9,
ymin=-5,
ymax=5,
xtick={-1.570796327,1.570796327},
xticklabels={\(-\tfrac{\pi}{2}\),\(\tfrac{\pi}{2}\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=200,
clip=false
]
\addplot[dashed][red] coordinates {(-1.570796327,-5) (-1.570796327,5)};
\addplot[dashed][red] coordinates {(1.570796327,-5) (1.570796327,5)};
\addplot[thick,blue, domain=-1.35:1.35] {tan(deg(x))};
\node at (axis cs:0.65,1.55) {\(\tan x\)};
\end{axis}
\end{tikzpicture}
\caption{Restricted branch of the tangent function used to define \(\arctan x\).}
\label{fig:restricted-tangent}
\end{figure}

\begin{definition}
The \emph{inverse tangent function}\index{inverse tangent function}\index{arctangent@$\arctan$}, denoted \(\arctan x\) or \(\tan^{-1} x\), is defined by
\[
\arctan x = y \quad \Longleftrightarrow \quad \tan y = x \text{ and } -\tfrac{\pi}{2} < y < \tfrac{\pi}{2}.
\]
Its domain is \(\mathbb{R}\) and its range is \(\left(-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right)\).
\end{definition}

\begin{proposition}[Inverse tangent identities]
\begin{conditiondisplay}
\tan(\arctan x) &= x, \qquad & x \in \mathbb{R},\\
\arctan(\tan x) &= x, \qquad & -\tfrac{\pi}{2} < x < \tfrac{\pi}{2}.
\end{conditiondisplay}
\end{proposition}

\begin{workedexample}
\begin{example}
Simplify the expression \(\cos(\arctan x)\).
\end{example}

\begin{solution}
Let \(y = \arctan x\), so \(\tan y = x\) with \(-\tfrac{\pi}{2} < y < \tfrac{\pi}{2}\). We want \(\cos y\). Using the identity \(1 + \tan^{2} y = \sec^{2} y\),
\[
\sec^{2} y = 1 + x^{2}.
\]
Since \(y \in \left(-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right)\) we have \(\cos y > 0\) and hence \(\sec y > 0\), so \(\sec y = \sqrt{1 + x^{2}}\). Therefore
\[
\cos(\arctan x) = \cos y = \frac{1}{\sec y} = \frac{1}{\sqrt{1 + x^{2}}}. \qedhere
\]
\end{solution}
\end{workedexample}

\subsection{The remaining inverse trigonometric functions}

The remaining inverse trigonometric functions are used less frequently, but they are defined in the same spirit.

\begin{definition}
The \emph{inverse cosecant function}\index{inverse cosecant function}\index{arccosecant@$\arccsc$}, denoted \(\arccsc x\) or \(\csc^{-1} x\), is defined by
\[
\arccsc x = y \quad \Longleftrightarrow \quad \csc y = x \text{ and } y \in \left(0, \tfrac{\pi}{2}\right] \cup \left(\pi, \tfrac{3\pi}{2}\right].
\]
\end{definition}

\begin{definition}
The \emph{inverse secant function}\index{inverse secant function}\index{arcsecant@$\arcsec$}, denoted \(\arcsec x\) or \(\sec^{-1} x\), is defined by
\[
\arcsec x = y \quad \Longleftrightarrow \quad \sec y = x \text{ and } y \in \left[0, \tfrac{\pi}{2}\right) \cup \left[\pi, \tfrac{3\pi}{2}\right).
\]
\end{definition}

\begin{definition}
The \emph{inverse cotangent function}\index{inverse cotangent function}\index{arccotangent@$\arccot$}, denoted \(\arccot x\) or \(\cot^{-1} x\), is defined by
\[
\arccot x = y \quad \Longleftrightarrow \quad \cot y = x \text{ and } y \in (0, \pi).
\]
\end{definition}

The domains of \(\arccsc\) and \(\arcsec\) are \((-\infty, -1] \cup [1, \infty)\), while the domain of \(\arccot\) is \(\mathbb{R}\). The principal ranges are the intervals specified in the definitions above.

\begin{remark}
For \(\arcsec x\), another common choice of principal range is \(y \in \left[0, \tfrac{\pi}{2}\right) \cup \left(\tfrac{\pi}{2}, \pi\right]\). In this text we use \(y \in \left[0, \tfrac{\pi}{2}\right) \cup \left[\pi, \tfrac{3\pi}{2}\right)\) because it makes the differentiation formulas for the inverse secant simpler. When consulting other sources or software, always check which convention is in use, since different branch choices can lead to different values.
\end{remark}

% expansion:intuition — parallels the sec^{-1} alternative-branch discussion so the csc^{-1} coverage is symmetric
\begin{remark}
For \(\arccsc x\), another common choice of principal range is \(y \in \left[-\tfrac{\pi}{2}, 0\right) \cup \left(0, \tfrac{\pi}{2}\right]\). In this text we use \(y \in \left(0, \tfrac{\pi}{2}\right] \cup \left(\pi, \tfrac{3\pi}{2}\right]\) so that the chosen branch parallels the one selected for \(\arcsec x\). Once the principal range is fixed consistently, either convention gives a valid inverse function.
\end{remark}

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.2, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

\section{The Limit of a Function}

% expansion:application — fleshes out the manuscript's shorthand "Ex: speed" into a motivated physical setup
A frequent setting in calculus is the need to study a ratio of changes. If \(s(t)\) is the position of a moving object at time \(t\), the ratio
\[
\frac{s(t + h) - s(t)}{h}
\]
is the \emph{average velocity}\index{average velocity} of the object from time \(t\) to time \(t + h\). Very often we want the \emph{instantaneous velocity}\index{instantaneous velocity} --- how fast the object is moving at a single moment --- rather than the average. To describe this, we let \(h\) shrink toward zero and ask what value the ratio approaches. More generally, for any function \(f\), the ratio
\[
\frac{f(y) - f(x)}{y - x}
\]
measures how \(f\) changes between two inputs. We may want to know what this ratio does as \(y\) gets closer and closer to \(x\), without ever equalling \(x\). This is the motivation for the concept of a limit.

\begin{definition}\label{def:limit-informal}
Suppose that \(f(x)\) is defined for all \(x\) near a number \(a\), except possibly at \(a\) itself. We write
\[
\lim_{x \to a} f(x) = L
\]
and say that the \emph{limit}\index{limit}\index{limit@$\lim$} of \(f(x)\) as \(x\) approaches \(a\) is \(L\) if \(f(x)\) can be made arbitrarily close to \(L\) whenever \(x\) is sufficiently close to \(a\) but not equal to \(a\).
\end{definition}

An alternative notation for the same idea is
\[
f(x) \to L \qquad \text{as } x \to a,
\]
read ``\(f(x)\) tends to \(L\) as \(x\) tends to \(a\).''

When we consider \(\lim_{x \to a} f(x)\), we are concerned with the values of \(f(x)\) for \(x\) near \(a\), not with the value of \(f(a)\). In particular, \(f(a)\) need not be defined at all. \Cref{fig:limit-same-near-a} shows three functions whose values at \(a\) are different (and one is undefined), yet which all share the same limit as \(x \to a\).

\begin{figure}[H]
\centering
\begin{minipage}[t]{0.32\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
    width=\linewidth,
    height=5cm,
    axis lines=middle,
    xmin=0, xmax=3.7,
    ymin=0, ymax=14,
    xtick=\empty,
    ytick=\empty,
    xlabel={\(x\)},
    ylabel={\(y\)},
    xlabel style={at={(axis description cs:0.98,0.03)},anchor=west},
    ylabel style={at={(axis description cs:0.06,0.98)},anchor=south},
    samples=200,
    clip=false
]
    \addplot[thick, blue, domain=0.45:3.45] {pow(2,x)};
    \addplot[dashed,red] coordinates {(0,8) (3,8)};
    \addplot[
        only marks,
        mark=*,
        mark size=2.6pt
    ] coordinates {(3,8)};
    \node[below] at (axis cs:3,0) {\(a\)};
    \node[left] at (axis cs:0,8) {\(L\)};
\end{axis}
\end{tikzpicture}
\vspace{0.3em}
\(f(a) = L\)
\end{minipage}\hfill
\begin{minipage}[t]{0.32\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
    width=\linewidth,
    height=5cm,
    axis lines=middle,
    xmin=0, xmax=3.7,
    ymin=0, ymax=14,
    xtick=\empty,
    ytick=\empty,
    xlabel={\(x\)},
    ylabel={\(y\)},
    xlabel style={at={(axis description cs:0.98,0.03)},anchor=west},
    ylabel style={at={(axis description cs:0.06,0.98)},anchor=south},
    samples=200,
    clip=false
]
    \addplot[thick, blue, domain=0.45:2.95] {pow(2,x)};
    \addplot[thick, blue, domain=3.05:3.45] {pow(2,x)};

    \addplot[dashed,red] coordinates {(0,8) (3,8)};

    \addplot[
        only marks,
        mark=o,
        mark size=2.6pt,
        mark options={draw=black, fill=white}
    ] coordinates {(3,8)};

    \addplot[
        only marks,
        mark=*,
        mark size=2.6pt
    ] coordinates {(3,5)};

    \node[below] at (axis cs:3,0) {\(a\)};
    \node[left] at (axis cs:0,8) {\(L\)};
\end{axis}
\end{tikzpicture}
\vspace{0.3em}
\(f(a) \neq L\)
\end{minipage}\hfill
\begin{minipage}[t]{0.32\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
    width=\linewidth,
    height=5cm,
    axis lines=middle,
    xmin=0, xmax=3.7,
    ymin=0, ymax=14,
    xtick=\empty,
    ytick=\empty,
    xlabel={\(x\)},
    ylabel={\(y\)},
    xlabel style={at={(axis description cs:0.98,0.03)},anchor=west},
    ylabel style={at={(axis description cs:0.06,0.98)},anchor=south},
    samples=200,
    clip=false
]
    \addplot[thick, blue, domain=0.45:2.95] {pow(2,x)};
    \addplot[thick, blue, domain=3.05:3.45] {pow(2,x)};

    \addplot[dashed,red] coordinates {(0,8) (3,8)};

    \addplot[
        only marks,
        mark=o,
        mark size=2.6pt,
        mark options={draw=black, fill=white}
    ] coordinates {(3,8)};

    \node[below] at (axis cs:3,0) {\(a\)};
    \node[left] at (axis cs:0,8) {\(L\)};
\end{axis}
\end{tikzpicture}
\vspace{0.3em}
\(f(a)\) is undefined
\end{minipage}

\caption{Three functions with the same limit as \(x \to a\).}
\label{fig:limit-same-near-a}
\end{figure}

\begin{workedexample}
\begin{example}\index{1/x near 0@$1/x$ near $0$}
Guess the value of
\[
\lim_{x \to 1} \frac{x - 1}{x^{2} - 1}.
\]
\end{example}

\begin{solution}
The expression is not defined at \(x = 1\), but we can examine its values for \(x\) near \(1\):
\[
\begin{array}{c cccccc}
\toprule
x & 0.9 & 0.99 & 0.999 & 1.001 & 1.01 & 1.1 \\
\midrule
\tfrac{x - 1}{x^{2} - 1} & 0.5263 & 0.5025 & 0.5003 & 0.4998 & 0.4975 & 0.4762 \\
\bottomrule
\end{array}
\]
The values are getting close to \(\tfrac{1}{2}\) as \(x\) approaches \(1\), so we guess
\[
\lim_{x \to 1} \frac{x - 1}{x^{2} - 1} = \tfrac{1}{2}. \qedhere
\]
\end{solution}
\end{workedexample}

\begin{workedexample}
\begin{example}
Guess the value of
\[
\lim_{t \to 0} \frac{\sqrt{t^{2} + 9} - 3}{t^{2}}.
\]
\end{example}

\begin{solution}
We evaluate the expression for values of \(t\) close to \(0\):
\[
\begin{array}{c cccc}
\toprule
t & -0.1 & -0.01 & 0.01 & 0.1 \\
\midrule
\tfrac{\sqrt{t^{2} + 9} - 3}{t^{2}} & 0.166620 & 0.166666 & 0.166666 & 0.166620 \\
\bottomrule
\end{array}
\]
These values are close to \(\tfrac{1}{6} = 0.166666\dots\), so we guess
\[
\lim_{t \to 0} \frac{\sqrt{t^{2} + 9} - 3}{t^{2}} = \tfrac{1}{6}. \qedhere
\]
\end{solution}
\end{workedexample}

The limits above were guessed by looking at function values near the point of interest. In the following sections, we will develop systematic methods for evaluating limits and study more subtle behavior such as one-sided limits and infinite limits.

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.3, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

\section{One-Sided and Infinite Limits}

% expansion:intuition — section opener flagging the two extensions of the basic limit concept
Near a point \(a\), a function can behave differently depending on which side \(x\) approaches from, and in some cases the values of the function do not approach any finite number at all. These two observations lead to one-sided limits and infinite limits.

\subsection{One-sided limits}

Sometimes we only want to consider one side of the approach to \(a\) --- for instance because the function is defined by different formulas on the two sides, or because the graph jumps at \(a\).

\begin{definition}
We write
\[
\lim_{x \to a^{-}} f(x) = L
\]
and say that the \emph{left-hand limit}\index{left-hand limit}\index{limit!left-hand} of \(f(x)\) as \(x\) approaches \(a\) is \(L\) if \(f(x)\) approaches \(L\) as \(x\) approaches \(a\) through values less than \(a\). Similarly,
\[
\lim_{x \to a^{+}} f(x) = L
\]
is the \emph{right-hand limit}\index{right-hand limit}\index{limit!right-hand}, describing the approach through values greater than \(a\).
\end{definition}

% expansion:figure — illustrative figure the manuscript implies with its one-sided discussion but does not draw
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=8.8cm,
height=5.4cm,
axis lines=middle,
xmin=-0.5,
xmax=4.5,
ymin=-0.5,
ymax=3.2,
xtick={2},
xticklabels={\(a\)},
ytick={1,2},
yticklabels={\(L_{1}\),\(L_{2}\)},
xlabel={\(x\)},
ylabel={\(y\)},
samples=120,
clip=false
]
\addplot[thick, blue, domain=0.2:1.95] {1 + 0.25*(x-2)^2};
\addplot[thick, blue, domain=2.05:4.2] {2 + 0.2*(x-2)^2};
\addplot[only marks, mark=o, mark size=2.2pt] coordinates {(2,1) (2,2)};
\end{axis}
\end{tikzpicture}
\caption{A function whose left-hand and right-hand limits at \(a\) are different.}
\label{fig:one-sided-limits}
\end{figure}

The two one-sided limits together determine the two-sided limit.

\begin{proposition}[Criterion for a two-sided limit]
\[
\lim_{x \to a} f(x) = L \quad \Longleftrightarrow \quad \lim_{x \to a^{-}} f(x) = L \text{ and } \lim_{x \to a^{+}} f(x) = L.
\]
\end{proposition}

If the two one-sided limits differ, then the two-sided limit does not exist.

% expansion:example — piecewise worked example illustrating the two-sided criterion in action; the manuscript introduces one-sided limits through definitions alone and this example makes the mechanism concrete
\begin{workedexample}
\begin{example}
Let
\[
f(x) =
\begin{cases}
x + 1, & x < 2,\\[4pt]
4 - x, & x \ge 2.
\end{cases}
\]
Evaluate \(\displaystyle\lim_{x \to 2^{-}} f(x)\) and \(\displaystyle\lim_{x \to 2^{+}} f(x)\), and determine whether \(\displaystyle\lim_{x \to 2} f(x)\) exists.
\end{example}

\begin{solution}
From the left we use the branch \(f(x) = x + 1\), so \(\displaystyle\lim_{x \to 2^{-}} f(x) = 3\).

From the right we use the branch \(f(x) = 4 - x\), so \(\displaystyle\lim_{x \to 2^{+}} f(x) = 2\).

Since the left-hand and right-hand limits differ, \(\displaystyle\lim_{x \to 2} f(x)\) does not exist.
\end{solution}
\end{workedexample}

\subsection{Infinite limits}

In some cases, the values of \(f(x)\) do not approach any finite number as \(x \to a\). Instead, they grow arbitrarily large positive or arbitrarily large negative.

\begin{workedexample}
\begin{example}
Consider \(f(x) = \dfrac{1}{x^{2}}\). What happens to \(f(x)\) as \(x \to 0\)?
\end{example}

\begin{solution}
If \(x\) is close to \(0\), then \(x^{2}\) is a very small positive number, so \(\dfrac{1}{x^{2}}\) is very large. For instance,
\[
\frac{1}{(0.1)^{2}} = 100, \qquad \frac{1}{(0.01)^{2}} = 10{,}000.
\]
The values of \(\dfrac{1}{x^{2}}\) grow without bound as \(x \to 0\), so no finite \(L\) satisfies the informal definition of the limit; the limit does not exist.
\end{solution}
\end{workedexample}

To capture this behavior, we introduce the notation
\[
\lim_{x \to 0} \frac{1}{x^{2}} = \infty.
\]
The symbol \(\infty\) does not represent a real number; it records that the function's values grow without bound.

\begin{definition}\label{def:infinite-limit-informal}
Let \(f\) be defined on both sides of \(a\), except possibly at \(a\) itself. We write
\[
\lim_{x \to a} f(x) = \infty
\]
if the values of \(f(x)\) can be made arbitrarily large by taking \(x\) sufficiently close to \(a\), with \(x \neq a\). We call such a limit an \emph{infinite limit}\index{infinite limit}\index{limit!infinite}. Similarly,
\[
\lim_{x \to a} f(x) = -\infty
\]
if the values of \(f(x)\) can be made arbitrarily large negative by taking \(x\) sufficiently close to \(a\), with \(x \neq a\).
\end{definition}

One-sided versions,
\[
\lim_{x \to a^{-}} f(x) = \infty,\qquad \lim_{x \to a^{+}} f(x) = \infty,\qquad \lim_{x \to a^{-}} f(x) = -\infty,\qquad \lim_{x \to a^{+}} f(x) = -\infty,
\]
are defined analogously. The four cases are illustrated in \cref{fig:one-sided-infinite-limits}.

\begin{figure}[H]
\centering

\begin{minipage}{0.23\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
width=\textwidth,
height=4.3cm,
axis lines=middle,
xmin=-0.2,
xmax=3.2,
ymin=-0.55,
ymax=3.2,
xtick={2},
xticklabels={\(a\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=120,
clip=false
]
\addplot[red, dashed] coordinates {(2.0,0) (2,3.0)};
\addplot[thick, blue, domain=0.25:1.68] {1/(2-x)};
\end{axis}
\end{tikzpicture}

\vspace{0.3em}
\[
\lim_{x \to a^{-}} f(x) = \infty
\]
\end{minipage}
\hfill
\begin{minipage}{0.23\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
width=\textwidth,
height=4.3cm,
axis lines=middle,
xmin=-0.2,
xmax=3.2,
ymin=-0.55,
ymax=3.2,
xtick={2},
xticklabels={\(a\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=120,
clip=false
]
\addplot[red, dashed] coordinates {(2.0,0) (2,3.0)};
\addplot[thick, blue, domain=2.32:3.0] {1/(x-2)};
\end{axis}
\end{tikzpicture}

\vspace{0.3em}
\[
\lim_{x \to a^{+}} f(x) = \infty
\]
\end{minipage}
\hfill
\begin{minipage}{0.23\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
width=\textwidth,
height=4.3cm,
axis lines=middle,
xmin=-0.2,
xmax=3.2,
ymin=-3.0,
ymax=0.75,
xtick={2},
xticklabels={\(a\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=120,
clip=false
]
\addplot[red, dashed] coordinates {(2,-3.0) (2,0.0)};
\addplot[thick, blue, domain=0.25:1.68] {-1/(2-x)};
\end{axis}
\end{tikzpicture}

\vspace{0.3em}
\[
\lim_{x \to a^{-}} f(x) = -\infty
\]
\end{minipage}
\hfill
\begin{minipage}{0.23\textwidth}
\centering
\begin{tikzpicture}
\begin{axis}[
width=\textwidth,
height=4.3cm,
axis lines=middle,
xmin=-0.2,
xmax=3.2,
ymin=-3.0,
ymax=0.75,
xtick={2},
xticklabels={\(a\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=120,
clip=false
]
\addplot[red, dashed] coordinates {(2,-3.0) (2,0.0)};
\addplot[thick, blue, domain=2.32:3.0] {-1/(x-2)};
\end{axis}
\end{tikzpicture}

\vspace{0.3em}
\[
\lim_{x \to a^{+}} f(x) = -\infty
\]
\end{minipage}

\caption{The four typical one-sided infinite limits.}
\label{fig:one-sided-infinite-limits}
\end{figure}

\subsection{Vertical asymptotes}

Infinite limits are closely related to vertical asymptotes.

\begin{definition}
The vertical line \(x = a\) is called a \emph{vertical asymptote}\index{vertical asymptote}\index{asymptote!vertical} of the curve \(y = f(x)\) if at least one of the following holds:
\[
\lim_{x \to a^{-}} f(x) = \pm\infty \qquad \text{or} \qquad \lim_{x \to a^{+}} f(x) = \pm\infty.
\]
\end{definition}

\begin{workedexample}
\begin{example}
Does the curve \(y = \dfrac{2x}{x - 3}\) have a vertical asymptote?
\end{example}

\begin{solution}
The denominator is zero only at \(x = 3\), so \(x = 3\) is the only candidate. As \(x \to 3^{+}\), the denominator \(x - 3\) is a small positive number while the numerator \(2x\) is close to \(6\), so
\[
\lim_{x \to 3^{+}} \frac{2x}{x - 3} = \infty.
\]
As \(x \to 3^{-}\), the denominator is a small negative number while the numerator is still close to \(6\), so
\[
\lim_{x \to 3^{-}} \frac{2x}{x - 3} = -\infty.
\]
At least one one-sided limit is infinite, so the line \(x = 3\) is a vertical asymptote, as shown in \cref{fig:vertical-asymptote-example}.
\end{solution}
\end{workedexample}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
width=8.5cm,
height=5.4cm,
axis lines=middle,
xmin=-1,
xmax=7,
ymin=-66,
ymax=66,
xtick={3},
xticklabels={\(3\)},
ytick=\empty,
xlabel={\(x\)},
ylabel={\(y\)},
samples=200,
clip=false
]
\addplot[dashed, red] coordinates {(3,-66) (3,66)};
\addplot[thick, blue, domain=-0.8:2.9] {2*x/(x-3)};
\addplot[thick, blue, domain=3.1:6.8] {2*x/(x-3)};
\end{axis}
\end{tikzpicture}
\caption{The graph of \(y = \dfrac{2x}{x - 3}\) has a vertical asymptote at \(x = 3\).}
\label{fig:vertical-asymptote-example}
\end{figure}

\begin{workedexample}
\begin{example}
Guess the value of \(\displaystyle \lim_{x \to 0^{+}} \ln x\).
\end{example}

\begin{solution}
As \(x \to 0^{+}\), the values of \(\ln x\) decrease without bound, so
\[
\lim_{x \to 0^{+}} \ln x = -\infty.
\]
Therefore \(x = 0\) is a vertical asymptote of the graph of \(y = \ln x\).
\end{solution}
\end{workedexample}

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.4, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

\section{Limit Laws and Computational Techniques}

% expansion:intuition — section opener pivoting from the graph/table approach of §§1.3-1.4 to the algebraic machinery of this section, and flagging that rigorous proofs wait until §1.6
In the previous sections we used tables and graphs to guess the values of limits. We now record algebraic properties that let us compute limits more systematically. These properties can be justified rigorously once we have the precise definition of a limit (see the next section); for now, they provide a reliable tool-kit.

\subsection{The limit laws}

\begin{theorem}[Limit laws]\index{limit laws}
Suppose \(c\) is a constant and both \(\lim_{x \to a} f(x)\) and \(\lim_{x \to a} g(x)\) exist. Then
\begin{aligneddisplay}
\lim_{x \to a}\bigl(f(x) + g(x)\bigr) &= \lim_{x \to a} f(x) + \lim_{x \to a} g(x),\\
\lim_{x \to a}\bigl(f(x) - g(x)\bigr) &= \lim_{x \to a} f(x) - \lim_{x \to a} g(x),\\
\lim_{x \to a}\bigl(c \, f(x)\bigr) &= c \lim_{x \to a} f(x),\\
\lim_{x \to a}\bigl(f(x) g(x)\bigr) &= \left(\lim_{x \to a} f(x)\right)\!\left(\lim_{x \to a} g(x)\right).
\end{aligneddisplay}
If \(\lim_{x \to a} g(x) \neq 0\), then
\[
\lim_{x \to a} \frac{f(x)}{g(x)} = \frac{\lim_{x \to a} f(x)}{\lim_{x \to a} g(x)}.
\]
Moreover, for every positive integer \(n\),
\[
\lim_{x \to a} \bigl[f(x)\bigr]^{n} = \left[\lim_{x \to a} f(x)\right]^{n}.
\]
\end{theorem}

Verbally: the limit of a sum / difference / constant multiple / product / integer power is the corresponding combination of the limits, and the limit of a quotient is the quotient of the limits whenever the denominator's limit is not zero.

These laws will be proved rigorously in the next section once we have the \(\varepsilon\)-\(\delta\) definition; for now, they let us evaluate many limits directly.

\subsection{Direct substitution}

A first consequence of the limit laws is that polynomials and rational functions behave in the easiest way possible.

\begin{proposition}
For every positive integer \(n\),
\begin{aligneddisplay}
\lim_{x \to a} x &= a,\\
\lim_{x \to a} x^{n} &= a^{n}.
\end{aligneddisplay}
Consequently, if \(p(x) = a_{n} x^{n} + a_{n-1} x^{n - 1} + \dots + a_{0}\) is a polynomial, then
\[
\lim_{x \to a} p(x) = p(a).
\]
If \(r(x)\) is a rational function and \(r(a)\) is defined, then
\[
\lim_{x \to a} r(x) = r(a).
\]
\end{proposition}

For polynomials and rational functions where the denominator does not vanish at \(a\), direct substitution is therefore the fastest route to the limit.

\begin{workedexample}
\begin{example}
Evaluate
\[
\lim_{x \to 1} \frac{x^{3} + 2x^{2} - 1}{5 - 3x}.
\]
\end{example}

\begin{solution}
The numerator and denominator are polynomials, so their limits at \(x = 1\) are computed by direct substitution:
\begin{aligneddisplay}
\lim_{x \to 1}\bigl(x^{3} + 2x^{2} - 1\bigr) &= 1 + 2 - 1 = 2,\\
\lim_{x \to 1}(5 - 3x) &= 5 - 3 = 2.
\end{aligneddisplay}
The denominator's limit is nonzero, so the quotient law gives
\[
\lim_{x \to 1} \frac{x^{3} + 2x^{2} - 1}{5 - 3x} = \frac{2}{2} = 1. \qedhere
\]
\end{solution}
\end{workedexample}

\subsection{Algebraic simplification}

When direct substitution produces an indeterminate form such as \(\tfrac{0}{0}\), we first simplify the expression and then take the limit.

\begin{workedexample}
\begin{example}
Evaluate
\[
\lim_{x \to 1} \frac{x - 1}{x^{2} - 1}.
\]
\end{example}

\begin{solution}
Direct substitution gives \(\tfrac{0}{0}\), so we simplify:
\begin{aligneddisplay}
\frac{x - 1}{x^{2} - 1} &= \frac{x - 1}{(x - 1)(x + 1)}\\
&= \frac{1}{x + 1}, \qquad x \neq 1.
\end{aligneddisplay}
Therefore
\[
\lim_{x \to 1} \frac{x - 1}{x^{2} - 1} = \lim_{x \to 1} \frac{1}{x + 1} = \tfrac{1}{2}. \qedhere
\]
\end{solution}
\end{workedexample}

\begin{workedexample}
\begin{example}
Evaluate
\[
\lim_{t \to 0} \frac{\sqrt{t^{2} + 9} - 3}{t^{2}}.
\]
\end{example}

\begin{solution}
Direct substitution gives \(\tfrac{0}{0}\), so we rationalise the numerator:
\begin{aligneddisplay}
\frac{\sqrt{t^{2} + 9} - 3}{t^{2}}
\cdot
\frac{\sqrt{t^{2} + 9} + 3}{\sqrt{t^{2} + 9} + 3}
&= \frac{t^{2} + 9 - 9}{t^{2}\bigl(\sqrt{t^{2} + 9} + 3\bigr)}\\
&= \frac{1}{\sqrt{t^{2} + 9} + 3}, \qquad t \neq 0.
\end{aligneddisplay}
Therefore
\[
\lim_{t \to 0} \frac{\sqrt{t^{2} + 9} - 3}{t^{2}} = \lim_{t \to 0} \frac{1}{\sqrt{t^{2} + 9} + 3} = \tfrac{1}{6}. \qedhere
\]
\end{solution}
\end{workedexample}

\subsection{Limits of piecewise-defined functions}

Some limits are easiest to analyse by splitting into cases.

\begin{workedexample}
\begin{example}
Evaluate \(\displaystyle \lim_{x \to 0} |x|\).
\end{example}

\begin{solution}
Using the piecewise description
\[
|x| =
\begin{cases}
x, & x \ge 0,\\[4pt]
-x, & x < 0,
\end{cases}
\]
we compute
\[
\lim_{x \to 0^{+}} |x| = \lim_{x \to 0^{+}} x = 0,
\qquad
\lim_{x \to 0^{-}} |x| = \lim_{x \to 0^{-}} (-x) = 0.
\]
Both one-sided limits equal \(0\), so \(\displaystyle \lim_{x \to 0} |x| = 0\).
\end{solution}
\end{workedexample}

\begin{workedexample}
\begin{example}
The greatest integer function \([x]\) is the largest integer less than or equal to \(x\), so that \([4] = 4\), \([4.1] = 4\), and \([4.8] = 4\). Determine whether \(\displaystyle \lim_{x \to 3} [x]\) exists.
\end{example}

\begin{solution}
As \(x \to 3^{+}\), the values of \(x\) lie just above \(3\), so \([x] = 3\) and \(\displaystyle \lim_{x \to 3^{+}} [x] = 3\). As \(x \to 3^{-}\), the values of \(x\) lie just below \(3\), so \([x] = 2\) and \(\displaystyle \lim_{x \to 3^{-}} [x] = 2\). The two one-sided limits differ, so \(\displaystyle \lim_{x \to 3} [x]\) does not exist.
\end{solution}
\end{workedexample}

\subsection{The squeeze theorem}

The limit laws require that both constituent limits exist, but sometimes one of the factors has no limit at all. A useful technique is to trap the expression between two simpler functions whose limits agree.

\begin{theorem}[Squeeze theorem]\index{squeeze theorem}
Suppose that
\[
g(x) \le f(x) \le h(x)
\]
for all \(x\) near \(a\), except possibly at \(x = a\). If
\pairdisplay{\lim_{x \to a} g(x) = L,}{\lim_{x \to a} h(x) = L,}
then \(\displaystyle \lim_{x \to a} f(x) = L\).
\end{theorem}

\begin{workedexample}
\begin{example}\index{x squared sin(1/x) example@$x^{2}\sin(1/x)$ example}
Evaluate
\[
\lim_{x \to 0} x^{2} \sin\!\left(\frac{1}{x}\right).
\]
\end{example}

\begin{solution}
We cannot apply the product law because \(\displaystyle \lim_{x \to 0} \sin(1/x)\) does not exist. Instead, we use the bound
\[
-1 \le \sin\!\left(\frac{1}{x}\right) \le 1
\]
valid for all \(x \neq 0\). Multiplying by \(x^{2} \ge 0\),
\[
-x^{2} \le x^{2} \sin\!\left(\frac{1}{x}\right) \le x^{2}.
\]
Since \(\displaystyle \lim_{x \to 0}(-x^{2}) = 0 = \lim_{x \to 0} x^{2}\), the squeeze theorem yields
\[
\lim_{x \to 0} x^{2} \sin\!\left(\frac{1}{x}\right) = 0. \qedhere
\]
\end{solution}
\end{workedexample}

% expansion:example — supplementary squeeze-theorem example using cosine and a single factor of x; reinforces the technique on a slightly different shape
\begin{workedexample}
\begin{example}
Evaluate
\[
\lim_{x \to 0} x \cos\!\left(\frac{1}{x}\right).
\]
\end{example}

\begin{solution}
Since \(\left|\cos(1/x)\right| \le 1\) for all \(x \neq 0\),
\[
\left|x \cos\!\left(\frac{1}{x}\right)\right| \le |x|,
\]
equivalently
\[
-|x| \le x \cos\!\left(\frac{1}{x}\right) \le |x|.
\]
Since \(\displaystyle \lim_{x \to 0}(-|x|) = 0 = \lim_{x \to 0} |x|\), the squeeze theorem yields
\[
\lim_{x \to 0} x \cos\!\left(\frac{1}{x}\right) = 0. \qedhere
\]
\end{solution}
\end{workedexample}

% expansion:strategy — distils the method across the worked examples in this section
\begin{strategy}[Computing a limit \(\lim_{x \to a} f(x)\)]
\begin{enumerate}
    \item \emph{Try direct substitution first.} If \(f(a)\) is defined and the expression is a polynomial or rational function with a nonzero denominator at \(a\), then the limit equals \(f(a)\).
    \item \emph{If substitution gives an indeterminate form} such as \(\tfrac{0}{0}\), simplify algebraically: factor and cancel, rationalise a numerator or denominator, or combine fractions.
    \item \emph{For a piecewise function}, evaluate the two one-sided limits on the appropriate branches and check whether they agree.
    \item \emph{If the expression is trapped between two simpler ones} whose limits at \(a\) are known and equal, apply the squeeze theorem.
    \item \emph{If the values grow without bound}, the limit is \(\pm\infty\) and the line \(x = a\) is a vertical asymptote.
\end{enumerate}
\end{strategy}

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.5, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

\section{The Precise Definition of a Limit}

% expansion:history [source: standard calculus-textbook historical note; cf. Stewart 8e §2.4 and Rogawski 4e Ch 2 openings] — Newton/Leibniz framing and the nineteenth-century ε-δ resolution, both of which are standard textbook register for opening a rigorous-limit section
From the time of Newton and Leibniz, calculus was remarkably successful at solving problems in geometry, mechanics, and astronomy. For a long time, however, there was also controversy about its logical foundations. Early arguments often relied on infinitesimals --- quantities thought of as being extremely small but not described in a fully precise way --- and mathematicians continued to ask what it really means to say that \(x\) approaches \(a\) or that \(f(x)\) approaches \(L\). The modern resolution, developed in the nineteenth century, is the \(\varepsilon\)-\(\delta\) definition\index{epsilon-delta definition@$\varepsilon$-$\delta$ definition} of a limit, which gave calculus the level of rigor used in modern mathematics.

In the previous sections we used graphs, tables, and algebraic techniques to study limits, and we stated the limit laws without proof. Those arguments rested on informal phrases such as ``\(f(x)\) is close to \(L\)'' and ``\(x\) is close to \(a\).'' To make the arguments rigorous --- and, eventually, to prove the limit laws --- we need a definition that replaces ``close'' with numerical precision.

To motivate the definition, consider the function
\[
f(x) =
\begin{cases}
2x - 1, & x \neq 3,\\[4pt]
6, & x = 3.
\end{cases}
\]
When \(x\) is close to \(3\) but \(x \neq 3\), the value \(f(x) = 2x - 1\) is close to \(5\). Intuitively, we expect \(\displaystyle \lim_{x \to 3} f(x) = 5\). But how close, exactly? We begin by making the informal phrase ``close'' numerically precise.

\begin{workedexample}
\begin{example}
For the function above, how close must \(x\) be to \(3\) so that \(f(x)\) differs from \(5\) by less than \(0.1\)?
\end{example}

\begin{solution}
We want \(|f(x) - 5| < 0.1\). For \(x \neq 3\),
\begin{aligneddisplay}
|f(x) - 5| &= |(2x - 1) - 5|\\
&= |2x - 6|\\
&= 2 |x - 3|.
\end{aligneddisplay}
Requiring \(2 |x - 3| < 0.1\) is equivalent to \(|x - 3| < 0.05\). Therefore, if \(0 < |x - 3| < 0.05\), then \(|f(x) - 5| < 0.1\).
\end{solution}
\end{workedexample}

Tightening the tolerance from \(0.1\) to \(0.01\), the same reasoning gives \(|f(x) - 5| < 0.01\) whenever \(0 < |x - 3| < 0.005\). As the tolerance shrinks, so must the neighborhood of \(3\). More generally, for any \(\varepsilon > 0\), the same manipulation gives \(|f(x) - 5| < \varepsilon\) whenever \(|x - 3| < \tfrac{\varepsilon}{2}\); we may therefore take \(\delta = \tfrac{\varepsilon}{2}\). The word \emph{arbitrarily} in the informal definition means precisely: we can answer the question for \emph{every} \(\varepsilon > 0\), not just for a few specific tolerances.

\begin{definition}\label{def:limit-precise}
Let \(f\) be a function defined on an open interval containing \(a\), except possibly at \(a\) itself. We say that
\[
\lim_{x \to a} f(x) = L
\]
if for every \(\varepsilon > 0\) there exists a \(\delta > 0\) such that
\[
|f(x) - L| < \varepsilon \qquad \text{whenever} \qquad 0 < |x - a| < \delta.
\]
\end{definition}

\Cref{fig:precise-limit} gives a geometric reading of the definition. If \(x\) is restricted to the punctured interval \((a - \delta, a + \delta)\) with \(x \neq a\), then the corresponding values of \(f(x)\) lie in the target strip \((L - \varepsilon, L + \varepsilon)\). The value \(f(a)\) itself may be different from \(L\): the limit depends only on the behavior of \(f(x)\) when \(x\) is close to \(a\), not on what happens at \(x = a\).

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=9cm,
    height=5.8cm,
    axis lines=middle,
    xmin=0, xmax=3.8,
    ymin=0, ymax=14,
    xtick=\empty,
    ytick=\empty,
    xlabel={\(x\)},
    ylabel={\(y\)},
    xlabel style={at={(axis description cs:0.98,0.03)},anchor=west},
    ylabel style={at={(axis description cs:0.06,0.98)},anchor=south},
    samples=200,
    clip=false
]
    \addplot[thick, blue, domain=0.45:2.95] {pow(2,x)};
    \addplot[thick, blue, domain=3.05:3.45] {pow(2,x)};

    \addplot[dashed, red] coordinates {(0,6.498019) (3.8,6.498019)};
    \addplot[dashed, red] coordinates {(0,9.849155) (3.8,9.849155)};

    \addplot[dashed, red] coordinates {(2.7,0) (2.7,14)};
    \addplot[dashed, red] coordinates {(3.3,0) (3.3,14)};

    \addplot[
        only marks,
        mark=o,
        mark size=2.8pt,
        mark options={draw=black, fill=white}
    ] coordinates {(3,8)};

    \addplot[
        only marks,
        mark=*,
        mark size=2.8pt
    ] coordinates {(3,11.6)};

    \addplot[
        only marks,
        mark=*,
        mark size=1.5pt
    ] coordinates {(2.7,6.498019)};

    \addplot[
        only marks,
        mark=*,
        mark size=1.5pt
    ] coordinates {(3.3,9.849155)};

    \node[below] at (axis cs:2.7,0) {\(a - \delta\)};
    \node[below] at (axis cs:3,0) {\(a\)};
    \node[below] at (axis cs:3.3,0) {\(a + \delta\)};

    \node[left] at (axis cs:0,6.498019) {\(L - \varepsilon\)};
    \node[left] at (axis cs:0,8) {\(L\)};
    \node[left] at (axis cs:0,9.849155) {\(L + \varepsilon\)};

    \draw[->] (axis cs:2.7,0.35) -- (axis cs:3,0.35);
    \draw[->] (axis cs:3.3,0.35) -- (axis cs:3,0.35);

    \draw[->] (axis cs:0.05,9.849155) -- (axis cs:0.05,8);
    \draw[->] (axis cs:0.05,6.498019) -- (axis cs:0.05,8);
\end{axis}
\end{tikzpicture}
\caption{Geometric interpretation of the \(\varepsilon\)-\(\delta\) definition; \(\delta\) depends on \(\varepsilon\).}
\label{fig:precise-limit}
\end{figure}

The condition \(0 < |x - a| < \delta\) says that \(x\) lies within distance \(\delta\) of \(a\), but \(x \neq a\). The inequality \(|f(x) - L| < \varepsilon\) says that \(f(x)\) lies within distance \(\varepsilon\) of \(L\). The key phrase is ``for every \(\varepsilon > 0\)'': no matter how small a tolerance \(\varepsilon\) we demand, we must be able to find a corresponding \(\delta\).

% expansion:formula — short standard proof of limit uniqueness via the triangle inequality; illustrates how the ε-δ definition is used and gives the section a first theoretical payoff
A first payoff of the \(\varepsilon\)-\(\delta\) definition is that the value of a limit, when it exists, is unambiguous.

\begin{proposition}[Uniqueness of limits]
If \(\displaystyle\lim_{x \to a} f(x)\) exists, then it is unique.
\end{proposition}

\begin{proof}
Suppose that
\begin{aligneddisplay}
\lim_{x \to a} f(x) &= L,\\
\lim_{x \to a} f(x) &= M,
\end{aligneddisplay}
and that \(L \neq M\). Let \(\varepsilon = \tfrac{|L - M|}{2}\). By the definition of the limit, there exist \(\delta_{1}, \delta_{2} > 0\) such that
\[
|f(x) - L| < \varepsilon \qquad \text{whenever} \qquad 0 < |x - a| < \delta_{1},
\]
and
\[
|f(x) - M| < \varepsilon \qquad \text{whenever} \qquad 0 < |x - a| < \delta_{2}.
\]
If \(0 < |x - a| < \min\{\delta_{1}, \delta_{2}\}\), both inequalities hold. By the triangle inequality,
\begin{aligneddisplay}
|L - M|
&= |L - f(x) + f(x) - M|\\
&\le |L - f(x)| + |f(x) - M|\\
&< 2\varepsilon\\
&= |L - M|,
\end{aligneddisplay}
which is impossible. Therefore \(L = M\).
\end{proof}

% expansion:strategy — distils the method across the two worked examples that follow
\begin{strategy}[Verifying a limit from the \(\varepsilon\)-\(\delta\) definition]
To show that \(\displaystyle \lim_{x \to a} f(x) = L\):
\begin{enumerate}
    \item Start with a generic \(\varepsilon > 0\). The task is to find a \(\delta > 0\) (which may depend on \(\varepsilon\)) that makes the definition work.
    \item Work with the quantity \(|f(x) - L|\) and bound it above by an expression involving \(|x - a|\).
    \item Choose \(\delta\) so that the bound above is less than \(\varepsilon\) whenever \(|x - a| < \delta\). When the bound carries an auxiliary factor depending on \(x\), restrict \(x\) to a small neighborhood of \(a\) to bound that factor; then take \(\delta\) to be the minimum of the neighborhood radius and the \(\varepsilon\)-based bound.
    \item Conclude that \(0 < |x - a| < \delta\) implies \(|f(x) - L| < \varepsilon\).
\end{enumerate}
\end{strategy}

\begin{workedexample}
\begin{example}
Show from the definition that \(\displaystyle \lim_{x \to 3} (4x - 5) = 7\).
\end{example}

\begin{solution}
Given \(\varepsilon > 0\), we seek \(\delta > 0\) with
\[
|(4x - 5) - 7| < \varepsilon \qquad \text{whenever} \qquad 0 < |x - 3| < \delta.
\]
Since
\begin{aligneddisplay}
|(4x - 5) - 7| &= |4x - 12|\\
&= 4 |x - 3|,
\end{aligneddisplay}
it is enough to require \(4 |x - 3| < \varepsilon\), i.e.\ \(|x - 3| < \tfrac{\varepsilon}{4}\). We choose \(\delta = \tfrac{\varepsilon}{4}\); then whenever \(0 < |x - 3| < \delta\),
\begin{aligneddisplay}
|(4x - 5) - 7| &= 4 |x - 3|\\
&< 4 \cdot \tfrac{\varepsilon}{4}\\
&= \varepsilon.
\end{aligneddisplay}
Therefore \(\displaystyle \lim_{x \to 3} (4x - 5) = 7\).
\end{solution}
\end{workedexample}

\begin{workedexample}
\begin{example}
Show from the definition that \(\displaystyle \lim_{x \to 1} (x^{2} - 5x + 6) = 2\).
\end{example}

\begin{solution}
Given \(\varepsilon > 0\), we seek \(\delta > 0\) with
\[
|(x^{2} - 5x + 6) - 2| < \varepsilon \qquad \text{whenever} \qquad 0 < |x - 1| < \delta.
\]
We first simplify:
\begin{aligneddisplay}
|(x^{2} - 5x + 6) - 2| &= |x^{2} - 5x + 4|\\
&= |(x - 1)(x - 4)|.
\end{aligneddisplay}
The factor \(|x - 4|\) depends on \(x\). To control it, first assume \(|x - 1| < 1\). Then
\begin{aligneddisplay}
|x - 1| < 1 &\implies -1 < x - 1 < 1\\
&\implies 0 < x < 2\\
&\implies -4 < x - 4 < -2\\
&\implies |x - 4| < 4.
\end{aligneddisplay}
So whenever \(|x - 1| < 1\), we have \(|(x - 1)(x - 4)| < 4 |x - 1|\). To force this below \(\varepsilon\), we also require \(4 |x - 1| < \varepsilon\), i.e.\ \(|x - 1| < \tfrac{\varepsilon}{4}\). Choosing
\[
\delta = \min\!\left\{1, \tfrac{\varepsilon}{4}\right\},
\]
we get both inequalities at once, and so
\begin{aligneddisplay}
|(x^{2} - 5x + 6) - 2| &= |(x - 1)(x - 4)|\\
&< 4 |x - 1|\\
&< 4 \cdot \tfrac{\varepsilon}{4}\\
&= \varepsilon.
\end{aligneddisplay}
Therefore \(\displaystyle \lim_{x \to 1} (x^{2} - 5x + 6) = 2\).
\end{solution}
\end{workedexample}

The \(\varepsilon\)-\(\delta\) machinery also gives precise meaning to the informal definitions of infinite limits from \cref{def:infinite-limit-informal}.

\begin{definition}\label{def:infinite-limit-precise}
We write
\[
\lim_{x \to a} f(x) = \infty
\]
if for every positive number \(M\) there exists \(\delta > 0\) such that
\[
f(x) > M \qquad \text{whenever} \qquad 0 < |x - a| < \delta.
\]
Similarly,
\[
\lim_{x \to a} f(x) = -\infty
\]
if for every negative number \(N\) there exists \(\delta > 0\) such that
\[
f(x) < N \qquad \text{whenever} \qquad 0 < |x - a| < \delta.
\]
\end{definition}

Intuitively, \(M\) may be any large positive threshold (such as \(100\), \(100^{100}\), or \(1000^{1000}\)): no matter how large the threshold, \(f(x)\) eventually exceeds it once \(x\) is close enough to \(a\). Likewise, \(N\) may be any negative threshold (such as \(-100\), \(-100^{100}\), or \(-1000^{1000}\)), below which \(f(x)\) is eventually forced.

% expansion:formula — ties the preceding definition to a property that will appear throughout later chapters
Finally, the \(\varepsilon\)-\(\delta\) definition gives a precise meaning to a property of functions that recurs throughout later chapters. If \(f\) is defined at every point near \(a\), including at \(a\), and if
\[
\lim_{x \to a} f(x) = f(a),
\]
then \(f\) is said to be \emph{continuous at \(a\)}\index{continuous function}\index{continuity}. A systematic study of continuity belongs to a later chapter; we mention the term here because it is expressed directly in terms of limits.

% TODO: add \subsection*{Exercises} block with end-of-section problems for Section 1.6, following CONTENT_SPEC.md §14 and CONTENT_EXERCISES.md.

% expansion:summary — the chapter-end Summary is required by CONTENT_SPEC §4; the items are synthesised from the manuscript content
\section*{Summary}

\paragraph{Key definitions.}
\begin{itemize}
    \item \emph{One-to-one function} --- different inputs always produce different outputs.
    \item \emph{Inverse function} \(f^{-1}\) --- reverses the action of a one-to-one function \(f\).
    \item \emph{Inverse trigonometric functions} \(\arcsin\), \(\arccos\), \(\arctan\), \(\arccsc\), \(\arcsec\), \(\arccot\) --- defined by restricting each trigonometric function to a principal interval on which it is one-to-one.
    \item \emph{Limit} \(\lim_{x \to a} f(x) = L\) --- informally, \(f(x)\) is arbitrarily close to \(L\) whenever \(x\) is sufficiently close to but not equal to \(a\); precisely, for every \(\varepsilon > 0\) there exists \(\delta > 0\) such that \(|f(x) - L| < \varepsilon\) whenever \(0 < |x - a| < \delta\).
    \item \emph{One-sided limits} \(\lim_{x \to a^{-}} f(x)\) and \(\lim_{x \to a^{+}} f(x)\) --- restrict the approach to values less than or greater than \(a\), respectively.
    \item \emph{Infinite limit} \(\lim_{x \to a} f(x) = \pm\infty\) --- the values of \(f(x)\) grow without bound as \(x\) approaches \(a\).
    \item \emph{Vertical asymptote} at \(x = a\) --- at least one of the one-sided limits at \(a\) is infinite.
    \item \emph{Continuity at a point} --- \(\lim_{x \to a} f(x) = f(a)\); studied systematically in a later chapter.
\end{itemize}

\paragraph{Key theorems, propositions, and corollaries.}
\begin{itemize}
    \item \emph{Horizontal line test} --- a function is one-to-one if and only if no horizontal line meets its graph more than once.
    \item \emph{Existence of an inverse} --- a function has an inverse if and only if it is one-to-one.
    \item \emph{Inverse trigonometric identities} --- on the principal intervals, \(\sin(\arcsin x) = x\) and \(\arcsin(\sin x) = x\), with analogous identities for \(\arccos\) and \(\arctan\).
    \item \emph{Criterion for a two-sided limit} --- \(\lim_{x \to a} f(x) = L\) if and only if both one-sided limits at \(a\) equal \(L\).
    \item \emph{Limit laws} --- limits respect sums, differences, products, quotients with nonzero denominator limit, and integer powers.
    \item \emph{Squeeze theorem} --- if \(g(x) \le f(x) \le h(x)\) near \(a\) and \(g\) and \(h\) share a common limit \(L\) there, then so does \(f\).
\end{itemize}

\paragraph{Key formulas and identities.}
\begin{itemize}
    \item Inverse composition: \(f^{-1}(f(x)) = x\) for \(x\) in the domain of \(f\), and \(f(f^{-1}(y)) = y\) for \(y\) in the range of \(f\).
    \item Principal intervals of the inverse trigonometric functions: \(\arcsin x \in \left[-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right]\), \(\arccos x \in [0, \pi]\), \(\arctan x \in \left(-\tfrac{\pi}{2}, \tfrac{\pi}{2}\right)\).
    \item Direct substitution: for polynomials and rational functions with nonzero denominator at \(a\), \(\lim_{x \to a} p(x) = p(a)\).
    \item Two-sided criterion: \(\lim_{x \to a} f(x) = L\) exactly when \(\lim_{x \to a^{\pm}} f(x) = L\).
    \item Precise \(\varepsilon\)-\(\delta\) statement: \(\lim_{x \to a} f(x) = L\) means that for every \(\varepsilon > 0\) there exists \(\delta > 0\) such that \(|f(x) - L| < \varepsilon\) whenever \(0 < |x - a| < \delta\).
\end{itemize}
